Phenotypic evolution is more constrained in simpler food webs

### 2018-11-23 manuscript.tex
updated figures and created a very rough draft o the entire discussion.

\documentclass[11pt,]{article}
\usepackage{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Matthew A. Barbour\^{}\{1,2,\textbackslash{}ast\}; Christopher J. Greyson-Gaito\^{}\{1,3\}; Arezoo Sootodeh\^{}\{3\}; Brendan Locke\^{}\{4\}; Jordi Bascompte\^{}\{2\}},
            pdftitle={Phenotypic evolution is more constrained in simple food webs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

  \title{Phenotypic evolution is more constrained in simple food webs}
    \author{Matthew A. Barbour\(^{1,2,\ast}\) \\ Christopher J. Greyson-Gaito\(^{1,3}\) \\ Arezoo Sootodeh\(^{3}\) \\ Brendan Locke\(^{4}\) \\ Jordi Bascompte\(^{2}\)}
    \date{}

@@

Adaptive landscapes describe the selective constraints acting on
phenotypic evolution, but often treat the community context of evolving
populations as a ``black box''. In contrast, ecological networks provide
an explicit representation of the community context, yet often ignore
the role of natural selection. Global change is simplifying the
structure of ecological networks; however, we are currently in a poor
position to predict how these simplified communities will affect the
evolutionary potential of populations. Here, we integrate ecological
networks and adaptive landscapes to examine how changes in food-web
complexity shape evolutionary constraints. We conducted a field
experiment that manipulated the diversity of insect parasitoids
(food-web complexity) that were able to impose selection on an insect
herbivore. We then measured herbivore survival as a function of three
key phenotypic traits. We found that more traits were under selection in
simple vs.~complex food webs. The adaptive landscape was more neutral in
complex food webs because different parasitoid species impose different
selection pressures, minimizing relative fitness differences among
phenotypes. Our results suggest that phenotypic evolution becomes more
constrained in simplified food webs. This indicates that the
simplification of ecological communities may constrain the adaptive
potential of populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes, to phenotypes, to species (Wright 1931; Simpson 1944; Arnold,
Pfrender, and Jones 2001). More than a metaphor, the adaptive landscape
links quantitative genetic and phenotypic variation to evolution by
natural selection (Lande 1979; Stevan J. Arnold and Wade 1984a; Stevan
J. Arnold and Wade 1984b). Ecological interactions often play a key role
in shaping adaptive landscapes, as evidenced by the role of antagonistic
(Schluter 2000; Abrams 2000) and mutualistic (Bronstein, Alarcón, and
Geber 2006) interactions in driving evolutionary change. Although there
is clear evidence that pairwise interactions can shape the adaptive
landscape, we also know that most species interact with multiple species
in a community context (Darwin 1859). Understanding how the adaptive
landscape is shaped by community context represents a major frontier for
evolutionary ecologists (McPeek 2017; terHorst et al. 2018). Resolution
on this front is urgently needed though, given the rapid impacts of
climate change on ecological communities (Scheffers et al. 2016).

Ecological networks, such as a food web describing who eats whom,
provide an explicit representation of the community context. These
networks describe the interdependency of populations within an
ecological community, providing an effective framework for predicting
how the strength and organization of ecological interactions (network
structure) affect population and community dynamics (Bascompte and
Jordano 2014; McCann 2012). Nevertheless, these networks often lack
quantitative information on how ecological interactions influence
selection on phenotypic traits. Thus, it remains unclear how network
structure shapes adaptive landscapes of interacting populations. This
represents a major barrier in our current ability to predict how changes
in community context will affect phenotypic evolution.

Here, we integrate ecological networks and adaptive landscapes to
understand how community context constrains evolutionary change (Arnold
1992). Different aspects of evolutionary constraints can be inferred by
quantifying the slope and curvature of the adaptive landscape. For
example, the slope is determined by directional selection gradients
acting on each phenotypic trait and influences the trajectory of
evolutionary change (Lande 1979; Arnold 1992). Evolutionary trajectories
become more constrained with an increase in the number of traits under
selection, as this diminishes the number of optimal solutions (Arnold
2003). The curvature of the adaptive landscape can also constrain
evolution through its indirect effect on genetic constraints (Arnold
1992; Hansen and Houle 2008). Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits (Hansen and Houle 2008). In general,
genetic constraints will increase with the number of traits under
directional or stabilizing selection, as this will decrease the additive
genetic variance in those traits (Hansen and Houle 2008). Genetic
constraints may also increase with the number of trait combinations
under correlational selection, as this type of selection decreases the
evolutionary independence of traits (Hansen and Houle 2008). If we want
to predict how community context constrains evolutionary change, we must
understand how ecological networks shape the adaptive landscape.

To examine the effect of network structure on adaptive landscapes, we
conducted a field experiment that manipulated the complexity of trophic
interactions (hereafter food-web complexity) associated with an abundant
insect herbivore (\emph{Iteomyia salicisverruca}; Family
Cecidomyiidae)(fig. \ref{fig:Conceptual}). The larvae of this herbivore
induce tooth-shaped galls when they feed on the developing leaves of
willow trees (\emph{Salix} sp., Russo 2006). These galls protect larva
from attack by generalist predators (e.g.~ants, spiders), but they
suffer high mortaltiy from egg and larval parasitoids (Barbour et al.
2016). We manipulated food-web complexity by either excluding the guild
of larval parasitoids (simple food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (complex food web;
fig. \ref{fig:Conceptual}). Food-web complexity may influence the
adaptive landscape in a number of ways. For example, if each parasitoid
imposes directional selection on a different trait, then more complex
food webs may constrain evolution by increasing the number of traits
under selection. Alternatively, if parasitoids impose selection on
different values of a trait, then their selective effects would cancel
each other out in more complex food webs. To examine these different
possibilities (among others), we applied modern statistical methods to
quantify the effect of food-web complexity on the slope and curvature of
the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

We conducted our study within a four-year old common garden experiment
of coastal willow (\emph{Salix hookeriana}) located at Humboldt Bay
National Wildlife Refuge (HBNWR) (40 40'53``N, 124 12'4''W) near Loleta,
California, USA. This common garden consists of 26 different willow
genotypes that were collected from a single population of willows
growing around Humboldt Bay. Stem cuttings of each genotype (25
replicates per genotypes) were planted in a completely randomized design
in two hectares of a former cattle pasture at HBNWR. Willows at our
study site begin flowering in February and reach their peak growth in
early August. During this study, willows had reached 5 - 9m in height.
Further details on the genotyping and planting of the common garden are
available in Barbour et al. (2015).

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population (Barbour et al. 2015). For
the complex food-web treatment (eight replicates per genotype), we used
flagging tape to mark 14 galled leaves per plant (\textasciitilde{}30
larvae), allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that we compared
control and treatment galls with similar phenology when we collected
galls later in the season. For the simple food-web treatment, we
enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant
Prairie, WI, USA) to exclude three parasitoid species that attack during
larval development. This treatment did not exclude the egg parasitoid
\emph{Platygaster} sp., which attacks prior to gall initiation (larva
initiate gall development in Cecidomyiid midges: Gagné 1989). In late
August, we collected marked and bagged galls from each plant, placed
them into 30 mL vials and kept them in the lab for 4 months at room
temperature. We then opened galls under a dissecting scope and
determined whether larvae survived to pupation (our measure of fitness)
or were parasitized. Since we were interested in selection imposed by
parasitoids, we excluded unknown sources of mortality. For the food-web
treatment that excluded larval parasitoids (simple food web), we further
restricted our data by removing any incidental instances of parasitism
by a larval parasitoid. This represented less than 3\% of the
observations in this food-web treatment and allowed us to focus our
inferences of selection on those imposed by the egg parasitoid. Our
final dataset contains survival estimates for 1285 larvae from 613 galls
and 111 plants.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work (Barbour et al.
2016) and others work with gall midges (Weis, Price, and Lynch 1983;
Heath, Abbot, and Stireman 2018). First, we measured gall diameter as
the size of each gall chamber to the nearest 0.01 mm at its maximum
diameter (perpendicular to the direction of plant tissue growth). Our
previous work has shown that larger galls are associated with higher
survival (Barbour et al. 2016). Second, we measured clutch size by
counting the number of chambers in each gall (Weis, Price, and Lynch
1983; Heath, Abbot, and Stireman 2018). All larvae collected from the
same multi-chambered gall were scored with the same clutch size. Third,
we measured oviposition (egg laying) preference as the density of larvae
observed on a plant in an independent survey. We did this randomly
sampling five branches per tree and counting the number of individual
gall chambers (number of larvae). We then converted these counts to a
measure of larval density per 100 shoots by counting the number of
shoots on the last branch we sampled. All larvae collected from the same
plant were scored with the same oviposition preference. Measuring larval
densities on plants in the field is a common method for measuring
oviposition preference (Gripenberg et al. 2010); however, caution must
be taken in inferring `preference' as larval densities can be influenced
by processes other than preference (Singer 1986). Fortunately, a couple
of features of our study system suggest that larval densities may be a
good proxy for oviposition preference. For example, since our data comes
from a randomized placement of plant genotypes in a common garden, there
is no consistent bias in which plant genotypes females are exposed to
while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general (Hawkins, Cornell,
and Hochberg 1997); therefore, we do not expect any prior egg predation
to bias our estimates of observed larval densities.

@@

Our analyses consisted of three parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
phenotypic traits (\(z_i\)) of individuals \textemdash in each food-web
treatment. Second, we translated selection surfaces into the scale of
relative fitness (\(w\)) in order to calculate selection gradients.
Third, we used our estimates of selection gradients to characterize the
slope and curvature of the adaptive landscape. Note that inferring
adaptive landscapes from selection surfaces assumes that trait
distributions are multivariate normal (Lande and Arnold 1983). To
approximate this assumption, we log-transformed clutch size and
square-root transformed oviposition preference. We then scaled all
phenotypic traits (mean=0 and SD=1) in order to calculate standardized
selection gradients that were comparable across traits and with other
studies of natural selection.

\textbf{Selection surface}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_{z_i}\)), quadratic (\(\alpha_{z_i:z_i}\)),
and linear interactions (\(\alpha_{z_i:z_j}\)) between each trait. We
also allowed these trait-fitness relationships (\(\alpha\)) to vary
between food-web treatments. Note that to obtain valid estimates of
linear trait-fitness relationships, we removed nonlinear terms prior to
estimating linear relationships (Lande and Arnold 1983). Other
approaches have been advocated for approximating selection surfaces
(Schluter 1988); however, our approach enables us to calculate selection
gradients, and thus is more appropriate for approximating the adaptive
landscape (Arnold 2003). To account for the nonindependence of clutch
size (measured at gall level) and oviposition preference (measured at
plant level) as well as any independent effects of willow genotype on
larval survival, we modelled gall ID nested within plant ID nested
within genotype ID as random effects. Although statistical models with
random effects are not common in analyses of natural selection, we think
this is a mistake since modelling random effects can mitigate biased
estimates of selection due to environmental covariances between traits
and fitness (Rausher 1992). Since our end goal was to characterize the
relationship between mean trait values and mean fitness (adaptive
landscape), we assumed the mean value of our random effects
(i.e.~setting them to zero) when calculating trait-fitness
relationships. We then used parametric bootstrapping (1,000 replicates)
to estimate the effect of food-web treatment on larval survival as well
as trait-fitness relationships in each food-web treatment. To determine
whether trait-fitness relationships differed between food-web
treatments, we calculated the difference in bootstrapped replicates
between treatments.

\textbf{Selection gradients}: We used the method of Janzen and Stern
(1998) to translate trait-fitness relationships (\(\alpha\)) into the
scale of relative fitness in order to calculate directional
(\(\beta_{z_i}\)), quadratic (\(\gamma_{z_i:z_i}\)), and correlational
(\(\gamma_{z_i:z_j}\)) selection gradients. Briefly, this method
calculates the average gradient of selection surfaces by multiplying the
average of \(W(z)[1-W(z)]\) by each regression coefficient (e.g.
\(\alpha_{z_i}\), \(\alpha_{z_i:z_i}\), or \(\alpha_{z_i:z_j}\)). We
then divided this average gradient by mean fitness (\(\bar W\)) to put
it on the scale of relative fitness, and thus interpretable as a
selection gradient. We estimated selection gradients separately for each
food-web treatment. Note that we doubled all quadratic terms prior to
calculating selection gradients to put them on the same scale as
estimates of directional and correlational selection (Stinchcombe et al.
2008).

\textbf{Adaptive landscape}: We took advantage of existing theory that
translates estimates of selection gradients to the adaptive landscape
(Phillips and Arnold 1989; Arnold 2003). The shape of the adaptive
landscape can be described by both its slope and curvature, each of
which describe different aspects of constraints on phenotypic evolution.
Specifically, the slope corresponds to the column vector of directional
selection gradients, which in our study corresponds to:

\[\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} \]
This slope describes the selective constraints acting on the trajectory
of phenotypic evolution, which is governed by the `Lande equation',
\(\Delta \bar z=\text{G}\beta\) (Lande 1979). Quantitative predictions
of evolutionary trajectories requires knowledge of a population's
\textbf{G}-matrix; however, we can gain qualitative insight to
evolutionary constraints by counting the number of phenotypic traits
under directional selection (selective constraints, Arnold 1992).

The curvature (\textbf{C}) of the adaptive landscape is a matrix that is
determined by directional, quadratic, and correlational selection
gradients, such that
\boldmath\(\text{C}=\gamma - \beta \beta^\text{T}\), or more explicitly:

@@

Note that we ommitted the upper triangle of each matrix for clarity
since it is simply the reflection of the lower triangle. This curvature
matrix describes the indirect effects of selection on genetic
constraints within a generation
(\(\Delta \text{G}=\text{G}(\gamma -\beta \beta^{\text{T}})\text{G}\)).
For example, the sign of diagonal terms of the curvature matrix dictate
whether selection will increase (+), decrease (-), or cause no change
(0) in the additive genetic variance of a trait. Similarly, any nonzero
covariance terms (off-diagonal) are indicative of selection for trait
integration (less trait independence). Therefore, we can infer the
indirect effects of selection on genetic constraints by counting the
number of negative signs along the diagonal (decrease in additive
genetic variance) and the number of nonzero terms along the off-diagonal
(trait integration) of the curvature matrix.

@@

All analyses and visualizations were conducted in R (R Core Team 2018).
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population (Arnold, Pfrender,
and Jones 2001). We also plot mean larval survival on a natural log
scale to accurately reflect the shape of the adaptive landscape (Arnold
2003).

@@

\hypertarget{refs}{}
\hypertarget{ref-Abrams2000}{}
Abrams, Peter A. 2000. ``The Evolution of Predator-Prey Interactions:
Theory and Evidence.'' \emph{Annual Review of Ecology and Systematics}
31 (1): 79--105.
doi:\href{https://doi.org/10.1146/annurev.ecolsys.31.1.79}{10.1146/annurev.ecolsys.31.1.79}.

\hypertarget{ref-Arnold1992}{}
Arnold, Stevan J. 1992. ``Constraints on Phenotypic Evolution.''
\emph{The American Naturalist} 140. University of Chicago Press:
S85--S107.

\hypertarget{ref-Arnold2003}{}
Arnold, Stevan J. 2003. ``Performance Surfaces and Adaptive
Landscapes1.'' \emph{Integrative and Comparative Biology} 43 (3):
367--75.
doi:\href{https://doi.org/10.1093/icb/43.3.367}{10.1093/icb/43.3.367}.

>>>>

\documentclass[11pt,]{article}
\usepackage{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Matthew A. Barbour\^{}\{1,2,\textbackslash{}ast\}; Christopher J. Greyson-Gaito\^{}\{1,3\}; Arezoo Sootodeh\^{}\{3\}; Brendan Locke\^{}\{4\}; Jordi Bascompte\^{}\{2\}},
            pdftitle={Phenotypic evolution is more constrained in simpler food webs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

  \title{Phenotypic evolution is more constrained in simpler food webs}
    \author{Matthew A. Barbour\(^{1,2,\ast}\) \\ Christopher J. Greyson-Gaito\(^{1,3}\) \\ Arezoo Sootodeh\(^{3}\) \\ Brendan Locke\(^{4}\) \\ Jordi Bascompte\(^{2}\)}
    \date{}

@@

Global change is simplifying the structure of ecological networks;
however, we are currently in a poor position to predict how these
simplified communities will affect the evolutionary potential of
remaining populations. Theory on adaptive landscapes provides a
framework for predicting how selection constrains phenotypic evolution,
but often treat the community context of evolving populations as a
``black box''. Here, we integrate ecological networks and adaptive
landscapes to examine how changes in food-web complexity shape
evolutionary constraints. We conducted a field experiment that
manipulated the diversity of insect parasitoids (food-web complexity)
that were able to impose selection on an insect herbivore. We then
measured herbivore survival as a function of three key phenotypic
traits. We found that more traits were under selection in simpler
vs.~more complex food webs. The adaptive landscape was more neutral in
complex food webs because different parasitoid species impose different
selection pressures, minimizing relative fitness differences among
phenotypes. Our results suggest that phenotypic evolution becomes more
constrained in simplified food webs. This indicates that the
simplification of ecological communities may constrain the adaptive
potential of remaining populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding
how natural selection has shaped the evolution of biodiversity ---from
genes to phenotypes to species
\citep{Wright1931, Simpson1944, Arnold2001}. More than a metaphor, the
adaptive landscape links quantitative genetic and phenotypic variation
to evolution by natural selection
\citep{Lande1979, Arnold1984applications, Arnold1984theory}. Ecological
interactions often play a key role in shaping adaptive landscapes, as
evidenced by the role of antagonistic \citep{Schluter2000, Abrams2000}
and mutualistic \citep{Bronstein2006} interactions in driving
evolutionary change. Although there is clear evidence that pairwise
interactions can shape the adaptive landscape, we do not know how the
adaptive landscape is shaped by the community context
\citep{McPeek2017, terHorst2018}. Resolution on how the community
context shapes phenotypic evolution is urgently needed though, given the
rapid impacts of climate change on ecological communities
\citep{Scheffers2016}.

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how community
context constrains evolutionary change \citep{Arnold1992}. Different
aspects of evolutionary constraints can be inferred by quantifying the
slope and curvature of the adaptive landscape. For example, the slope is
determined by directional selection gradients acting on each phenotypic
trait and influences the trajectory of evolutionary change
\citep{Lande1979, Arnold1992}. Evolutionary trajectories become more
constrained with an increase in the number of traits under selection, as
this diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape can also constrain evolution through
its indirect effect on genetic constraints
\citep{Arnold1992, Hansen2008}. Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits \citep{Hansen2008}. In general, genetic
constraints will increase with the number of traits under directional or
stabilizing selection, as this will decrease the additive genetic
variance in those traits \citep{Hansen2008}. Genetic constraints may
also increase with the number of trait combinations under correlational
selection, as this type of selection decreases the evolutionary
independence of traits \citep{Hansen2008}. If we want to predict how
community context constrains evolutionary change, we must understand how
ecological networks shape the adaptive landscape.

Global change is simplifying the structure of ecological communities,
which may influence evolutionary constraints in a number of ways. For
example, in a multitrophic community or food web, if different consumers
impose directional selection on different traits of a shared resource,
then more complex food webs may constrain evolution by increasing the
number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more complex food webs. To examine these
different possibilities (among others), we conducted a field experiment
that manipulated the complexity of trophic interactions (hereafter
food-web complexity) associated with an abundant insect herbivore
(\emph{Iteomyia salicisverruca}; Family Cecidomyiidae)(fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortaltiy from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web complexity by either excluding the guild of larval
parasitoids (simple food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (complex food web;
fig. \ref{fig:Conceptual}). We then applied modern statistical methods
to quantify the effect of food-web complexity on the slope and curvature
of the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

We conducted our study within a four-year old common garden experiment
of coastal willow (\emph{Salix hookeriana}) located at Humboldt Bay
National Wildlife Refuge (HBNWR) (40 40'53``N, 124 12'4''W) near Loleta,
California, USA. This common garden consists of 26 different willow
genotypes that were collected from a single population of willows
growing around Humboldt Bay. Stem cuttings of each genotype (25
replicates per genotypes) were planted in a completely randomized design
in two hectares of a former cattle pasture at HBNWR. Willows at our
study site begin flowering in February and reach their peak growth in
early August. During this study, willows had reached 5 - 9m in height.
Further details on the genotyping and planting of the common garden are
available in \citet{Barbour2015}.

@@

We setup our food-web manipulation across 128 plants soon after galls
began developing on willows in early June of 2013. These 128 plants came
from eight different plant genotypes that spanned the range of trait
variation observed in this willow population \citep{Barbour2015}. For
the complex food-web treatment (eight replicates per genotype), we used
flagging tape to mark 14 galled leaves per plant (\textasciitilde{}30
larvae), allowing the full suite of egg and larval parasitoids to impose
selection. Marking galls with flagging tape ensured that we compared
control and treatment galls with similar phenology when we collected
galls later in the season. For the simple food-web treatment, we
enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant
Prairie, WI, USA) to exclude three parasitoid species that attack during
larval development. This treatment did not exclude the egg parasitoid
\emph{Platygaster} sp., which attacks prior to gall initiation
\citep[larva initiate gall development in Cecidomyiid
midges:][]{Gagne1989}. In late August, we collected marked and bagged
galls from each plant, placed them into 30 mL vials and kept them in the
lab for 4 months at room temperature. We then opened galls under a
dissecting scope and determined whether larvae survived to pupation (our
measure of fitness) or were parasitized. Since we were interested in
selection imposed by parasitoids, we excluded unknown sources of
mortality. For the food-web treatment that excluded larval parasitoids
(simple food web), we further restricted our data by removing any
incidental instances of parasitism by a larval parasitoid. This
represented less than 3\% of the observations in this food-web treatment
and allowed us to focus our inferences of selection on those imposed by
the egg parasitoid. Our final dataset contains survival estimates for
1285 larvae from 613 galls and 111 plants.

@@

We collected data on three different traits that we expected to
influence larval survival based on our previous work \citep{Barbour2016}
and others work with gall midges \citep{Weis1983, Heath2018}. First, we
measured gall diameter as the size of each gall chamber to the nearest
0.01 mm at its maximum diameter (perpendicular to the direction of plant
tissue growth). Our previous work has shown that larger galls are
associated with higher survival \citep{Barbour2016}. Second, we measured
clutch size by counting the number of chambers in each gall
\citep{Weis1983, Heath2018}. All larvae collected from the same
multi-chambered gall were scored with the same clutch size. Third, we
measured oviposition preference as the density of larvae observed on a
plant in an independent survey. We did this randomly sampling five
branches per tree and counting the number of individual gall chambers
(number of larvae). We then converted these counts to a measure of
larval density per 100 shoots by counting the number of shoots on the
last branch we sampled. All larvae collected from the same plant were
scored with the same oviposition preference. Measuring larval densities
on plants in the field is a common method for measuring oviposition
preference \citep{Gripenberg2010}; however, caution must be taken in
inferring `preference' as larval densities can be influenced by
processes other than preference \citep{Singer1986}. Fortunately, a
couple of features of our study system suggest that larval densities may
be a good proxy for oviposition preference. For example, since our data
comes from a randomized placement of plant genotypes in a common garden,
there is no consistent bias in which plant genotypes females are exposed
to while searching for oviposition sites. Also, egg predation is a minor
source of mortality for galling insects in general \citep{Hawkins1997};
therefore, we do not expect any prior egg predation to bias our
estimates of observed larval densities.

@@

Our analyses consisted of three parts. First, we used generalized linear
mixed models (GLMM) to quantify selection surfaces \textemdash linear
and nonlinear relationships between absolute fitness (\(W\)) and
phenotypic traits (\(z_i\)) of individuals \textemdash in each food-web
treatment. Second, we translated selection surfaces into the scale of
relative fitness (\(w\)) in order to calculate selection gradients.
Third, we used our estimates of selection gradients to characterize the
slope and curvature of the adaptive landscape. Note that inferring
adaptive landscapes from selection surfaces assumes that trait
distributions are multivariate normal \citep{Lande1983}. To approximate
this assumption, we log-transformed clutch size and square-root
transformed oviposition preference. We then scaled all phenotypic traits
(mean=0 and SD=1) in order to calculate standardized selection gradients
that were comparable across traits and with other studies of natural
selection.

\textbf{Selection surface}: Since larval survival was our measure of
absolute fitness, we used a GLMM that assumed a binomial error
distribution (and logit-link function). To approximate the selection
surface, we modelled larval survival as a function of food-web treatment
as well as linear (\(\alpha_{z_i}\)), quadratic (\(\alpha_{z_i:z_i}\)),
and linear interactions (\(\alpha_{z_i:z_j}\)) between each trait. We
also allowed these trait-fitness relationships (\(\alpha\)) to vary
between food-web treatments. Note that to obtain valid estimates of
linear trait-fitness relationships, we removed nonlinear terms prior to
estimating linear relationships \citep{Lande1983}. Other approaches have
been advocated for approximating selection surfaces
\citep{Schluter1988}; however, our approach enables us to calculate
selection gradients, and thus is more appropriate for approximating the
adaptive landscape \citep{Arnold2003}. To account for the
nonindependence of clutch size (measured at gall level) and oviposition
preference (measured at plant level) as well as any independent effects
of willow genotype on larval survival, we modelled gall ID nested within
plant ID nested within genotype ID as random effects. Although
statistical models with random effects are not common in analyses of
natural selection, we think that modelling random effects can mitigate
biased estimates of selection due to environmental covariances between
traits and fitness \citep{Rausher1992}. Since our end goal was to
characterize the relationship between mean trait values and mean fitness
(adaptive landscape), we assumed the mean value of our random effects
(i.e., setting them to zero) when calculating trait-fitness
relationships. We then used parametric bootstrapping (1,000 replicates)
to estimate the effect of food-web treatment on larval survival as well
as trait-fitness relationships in each food-web treatment. To determine
whether trait-fitness relationships differed between food-web
treatments, we calculated the difference in bootstrapped replicates
between treatments.

\textbf{Selection gradients}: We used the method of \citet{Janzen1998}
to translate trait-fitness relationships (\(\alpha\)) into the scale of
relative fitness in order to calculate directional (\(\beta_{z_i}\)),
quadratic (\(\gamma_{z_i:z_i}\)), and correlational
(\(\gamma_{z_i:z_j}\)) selection gradients. Briefly, this method
calculates the average gradient of selection surfaces by multiplying the
average of \(W(z)[1-W(z)]\) by each regression coefficient (e.g.
\(\alpha_{z_i}\), \(\alpha_{z_i:z_i}\), or \(\alpha_{z_i:z_j}\)). We
then divided this average gradient by mean fitness (\(\bar W\)) to put
it on the scale of relative fitness, and thus interpretable as a
selection gradient. We estimated selection gradients separately for each
food-web treatment. Note that we doubled all quadratic terms prior to
calculating selection gradients to put them on the same scale as
estimates of directional and correlational selection
\citep{Stinchcombe2008}.

\textbf{Evolutionary constraints}: We quantified the effects of food-web
complexity on evolutionary constraints by inspecting the slope and
curvature of the adaptive landscape. The number of selective constraints
is determined by the slope of the adaptive landscape, which in our study
corresponds to:

\[\text{Slope} = \beta = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} \]
By comparing the number of traits that determine the slope (i.e.,
nonzero directional selection gradients) in simpler vs.~more complex
food webs, we can infer the effect of food-web complexity on selective
constraints.

The indirect effects of selection on genetic constraints is governed by
the curvature of the adaptive landscape, which in our study corresponds
to:

@@

Note that we ommitted the upper triangle of each matrix for clarity
since it is simply the reflection of the lower triangle. The sign of
diagonal terms of the curvature matrix dictate whether selection will
increase (+), decrease (-), or cause no change (0) in the additive
genetic variance of a trait. Similarly, any nonzero covariance terms
(off-diagonal) are indicative of selection for trait integration (i.e.,
less independence). Therefore, we can infer the indirect effects of
selection on genetic constraints by counting the number of negative
signs along the diagonal (decrease in additive genetic variance) and the
number of nonzero terms along the off-diagonal (trait integration) of
the curvature matrix.

@@

All analyses and visualizations were conducted in R \citep{R2018}.
Unless otherwise noted, we report mean estimates of trait-fitness
relationships and selection gradients with 95\% confidence intervals in
brackets. Note that for visualizing the adaptive landscape we restrict
trait axes to \(\pm 1\) SD of the mean trait value. This emphasizes the
fact that we can only reliably estimate the shape of the adaptive
landscape near the mean phenotype of the population \citep{Arnold2001}.
We also plot mean larval survival on a natural log scale to accurately
reflect the shape of the adaptive landscape \citep{Arnold2003}.

@@

\hypertarget{refs}{}
